---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706
### Lab03 - Data Preparation and Cleaning in R
#### Tutorial
*Alessandro Pandini*
Revision: 1.01

***
#### Table of contents

1. [Detecting missing values](#Detecting_missing_values)  
2. [Imputing missing values / removing incomplete instances](#Imputing_missing_values)  
3. [Removing duplicate instances](#Removing_duplicate_instances)  
4. [(Simple) outlier detection](#outlier_detection)  
5. [Subsetting a data frame](#Subsetting_data_frame)  
6. [Sampling instances from a data frame](#Sampling_instances)  

***

#### 1. Detecting missing values{#Detecting_missing_values}
```{r}
# read RMS Titanic data from a csv file
#   note: NA and empty strings are interpreted as NA values
RMS_Titanic <- read.csv("RMS_Titanic.csv", na.strings = c("NA", ""))
# inspect the data set
str(RMS_Titanic)

# convert the Survived and Pclass variable to factors
RMS_Titanic$Survived <- factor(RMS_Titanic$Survived)
RMS_Titanic$Pclass <- factor(RMS_Titanic$Pclass)
# inspect the dataset to see the changes you made
str(RMS_Titanic)

# NA (Not Available) is interpreted as 'missing value' in R
# get summary report of the variables to look for NAs
#   three variables (Age, Cabin, Embarked) have NA values
summary(RMS_Titanic)

# get a table of the number of NA per variable
RMS_Titanic_NA_count <- apply(is.na(RMS_Titanic), 2, sum)
RMS_Titanic_NA_count
# ...and the percentage of NA per variable
RMS_Titanic_NA_perc <- RMS_Titanic_NA_count / dim(RMS_Titanic)[1] * 100
RMS_Titanic_NA_perc
```

#### 2. Imputing missing values / removing incomplete instances{#Imputing_missing_values}
```{r}
# impute by replacement with the most frequent value (for Port of Embarkation)
#   first calculate the frequency of values (C, Q, S) for Embarked
#     S is the most frequent
table(RMS_Titanic$Embarked)
#   then set Embarked missing values to S
RMS_Titanic[is.na(RMS_Titanic$Embarked), 'Embarked'] = 'S'

# inspect the frequency of values for Embarked after imputation
table(RMS_Titanic$Embarked)

# impute missing Age values by replacement with the median value
#   first get the age median
#     note: NA should be removed to calculate the median!
median_age <- median(RMS_Titanic$Age, na.rm = T)
#   then set Age missing values to its median
RMS_Titanic[is.na(RMS_Titanic$Age), 'Age'] = median_age

# remove instances with missing values in all other attributes
RMS_Titanic_noNA <- na.omit(RMS_Titanic)

# check the size of the data frames before and after removal
dim(RMS_Titanic)
dim(RMS_Titanic_noNA)

# note: R functions often have an na.action option
#   this option defines how NAs will be treated in the analysis
#   it is particularly important, if NAs have not been removed from the data
```

#### 3. Removing duplicate instances{#Removing_duplicate_instances}
```{r}
# remove duplicate rows
RMS_Titanic_clean <- unique(RMS_Titanic_noNA)

# check the size of the data frames before and after removal
dim(RMS_Titanic_noNA)
dim(RMS_Titanic_clean)

# find the duplicate rows
#   duplicated returns a boolean vector
RMS_Titanic_duplicated <- duplicated(RMS_Titanic_noNA)
RMS_Titanic_duplicated

# duplicated rows can then be extracted by conditional filtering
RMS_Titanic_noNA[RMS_Titanic_duplicated, ]
```

#### 4. (Simple) outlier detection{#outlier_detection}
```{r}
# inspect the Fare distribution using summary statistics
summary(RMS_Titanic_clean$Fare)
# generate a boxplot of the Fare variable
boxplot(RMS_Titanic_clean$Fare)

# outliers can be detected from the boxplot
#   note: the boxplot command returns a summary statistics object
#     this object can be assigned to a variable and inspected
RMS_Titanic_boxplot <- boxplot(RMS_Titanic_clean$Fare)
# inspect the boxplot statistics
#   note: the returened value is a list object
RMS_Titanic_boxplot
# outlier values can be extracted from this object
#   they are stored in the element 'out' of the list
RMS_Titanic_boxplot$out

# get the minimum value that is an outlier
min(RMS_Titanic_boxplot$out)

# outliers rows can be extracted by conditional selection
RMS_Titanic_clean[RMS_Titanic_clean$Fare >= min(RMS_Titanic_boxplot$out), ]
```

#### 5. Subsetting a data frame{#Subsetting_data_frame}
```{r}
# subset the dataset for age < 10
RMS_Titanic_clean_10 <- subset(RMS_Titanic_clean, Age < 10)
RMS_Titanic_clean_10

# subset the dataset for age > 60 and Sex == male
RMS_Titanic_clean_60_male <- subset(RMS_Titanic_clean, Age > 60 & Sex == 'male')
RMS_Titanic_clean_60_male
```

#### 6. Sampling instances from a data frame}{#Sampling_instances}
```{r}
# random sample of 10 rows from the data frame
#   first sample 10 number from the row indices
sample_idx <- sample(nrow(RMS_Titanic_clean), 10)
sample_idx
#   then extract the instances using the sampled indices
RMS_Titanic_clean[sample_idx,]

# random sample of 100 indices w/ replacement
sample_idx <- sample(nrow(RMS_Titanic_clean), 100, replace = T)
# inspect the sample (there should be repeated values)
sample_idx

# create a 70/30 training/test set split
#   note: test set indices are derived as complement of the training ones
#   first get the number of rows in the data frame
n_rows <- nrow(RMS_Titanic_clean)
# sample 70% (n_rows * 0.7) indices in the ranges 1:nrows
training_idx <- sample(n_rows, n_rows * 0.7)
# filter the data frame with the training indices (and the complement)
training_RMS_Titanic_clean <- RMS_Titanic_clean[training_idx,]
test_RMS_Titanic_clean <- RMS_Titanic_clean[-training_idx,]

# check the size of the two data sets
dim(training_RMS_Titanic_clean)
dim(test_RMS_Titanic_clean)

# stratified sampling is available from the sampling package
# install the sampling package from CRAN
if(require(sampling) == FALSE){
    install.packages("sampling")
}

# stratified sampling by Sex
#   extract 8 rows from the 'female' partition and 5 from the 'male' one
sampling::strata(RMS_Titanic_clean, "Sex", c(8,5))

######################################################################
#
# for more advanced data frame handling see the package:
#   dplyr
#   https://cran.rstudio.com/web/packages/dplyr/vignettes/base.html
#
######################################################################
```

