---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706
### Lab 04 - Exploratory Data Analysis in R
#### Tutorial
*Alessandro Pandini*
Revision: 1.01

Reference:  
Chapter 11: Dimension Reduction
Machine Learning with R Cookbook
Yu-Wei, Chiu
Packt Publishing 2015  

***
#### Table of contents

1. [Summary statistics](#Summary_statistics)  
2. [Graphical analysis](#Graphical_analysis)  
3. [Principal Component Analysis](#Principal_Component_Analysis)  
4. [Visual analysis of PCA results](#Visual_analysis_PCA)  

***

#### 1. Summary statistics{#Summary_statistics}
```{r}
# load the swiss data set from R dataset package
data(swiss)
#   for detailed information about other available datasets:
#     library(help = "datasets")

# get the documentation about this dataset with the help command
#   note: all variables are in percentage, i.e. [0,100] range
help(swiss)

# inspect the dataset
str(swiss)

# get a summary report
summary(swiss)

View(swiss)
# calculate person correlation coefficient for each pair of variables
cor(swiss)
```

#### 2. Graphical analysis{#Graphical_analysis}
```{r}
# generate a histogram for each variable (and show them on the same page)
#   note: titles and x labels are set to the name of the relevant variable
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,3))
hist(swiss[, 1], main = names(swiss)[1], xlab = names(swiss)[1], xlim = c(0,100))
hist(swiss[, 2], main = names(swiss)[2], xlab = names(swiss)[2], xlim = c(0,100))
hist(swiss[, 3], main = names(swiss)[3], xlab = names(swiss)[3], xlim = c(0,100))
hist(swiss[, 4], main = names(swiss)[4], xlab = names(swiss)[4], xlim = c(0,100))
hist(swiss[, 5], main = names(swiss)[5], xlab = names(swiss)[5], xlim = c(0,100))
hist(swiss[, 6], main = names(swiss)[6], xlab = names(swiss)[6], xlim = c(0,100))
par(opar)

# generate a density plot for each variable (and show them on the same page)
#   note: kernel density estimation may have tails outside the variable range!
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,3))
plot(density(swiss[, 1]), main = names(swiss)[1], xlab = names(swiss)[1])
plot(density(swiss[, 2]), main = names(swiss)[2], xlab = names(swiss)[2])
plot(density(swiss[, 3]), main = names(swiss)[3], xlab = names(swiss)[3])
plot(density(swiss[, 4]), main = names(swiss)[4], xlab = names(swiss)[4])
plot(density(swiss[, 5]), main = names(swiss)[5], xlab = names(swiss)[5])
plot(density(swiss[, 6]), main = names(swiss)[6], xlab = names(swiss)[6])
par(opar)

# generate a boxplot graph including horizontal background dashed lines
#   note: this can be done by
#     1. plotting an empty graph, i.e. with white boxes
#     2. adding the background dashed lines
#     3. plotting the coloured boxplot with the option add = T
opar <- par(no.readonly = TRUE)
boxplot(
  swiss,
  border = 'white',
  yaxt = 'n'
)
abline(h = seq(0,100,10), lty = 'dashed', lwd = 2, col = 'darkgrey')
boxplot(
  swiss,
  border = 'blue',
  yaxt = 'n',
  add = T
)
axis(2, seq(0,100,10))
par(opar)
```

#### 3. Principal Component Analysis{#Principal_Component_Analysis}
```{r}
# perform PCA on the swiss dataset
#   note: variables are centered and scaled before analysis
pc_swiss <- prcomp(swiss, center = T, scale. = T)


# inspect the attributes of the PCA object returned by prcomp
attributes(pc_swiss)
# see value section of the help for the prcomp for more details
help(prcomp)

summary(prcomp)
```

#### 4. Visual analysis of PCA results{#Visual_analysis_PCA}
```{r}
# calculate the proportion of exaplained variance (PEV) from the std values
pc_swiss_var <- pc_swiss$sdev^2
pc_swiss_var
pc_swiss_PEV <- pc_swiss_var / sum(pc_swiss_var)
pc_swiss_PEV

# plot the variance per PC
#   note: this can be done using the plot function on the prcomp object
plot(pc_swiss)

# plot the cumulative value of PEV for increasing number of additional PCs
#   note: add an 80% threshold line to inform the feature extraction
#     according to the plot the first 3 PCs should be selected
opar <- par(no.readonly = TRUE)
plot(
  cumsum(pc_swiss_PEV),
  ylim = c(0,1),
  xlab = 'PC',
  ylab = 'cumulative PEV',
  pch = 20,
  col = 'orange'
)
abline(h = 0.8, col = 'red', lty = 'dashed')
par(opar)

# get and inspect the loadings for each PC
#   note: loadings are reported as a rotation matrix (see lecture)
pc_swiss_loadings <- pc_swiss$rotation
pc_swiss_loadings

# plot the loadings for the first three PCs as a barplot
#   note: two vectors for colours and labels are created for convenience
#     for details on the other parameters see the help for barplot and legend
opar <- par(no.readonly = TRUE)
colvector = c('red', 'orange', 'yellow', 'green', 'cyan', 'blue')
labvector = c('PC1', 'PC2', 'PC3')
barplot(
  pc_swiss_loadings[,c(1:3)],
  beside = T,
  yaxt = 'n',
  names.arg = labvector,
  col = colvector,
  ylim = c(-1,1),
  border = 'white',
  ylab = 'loadings'
)
axis(2, seq(-1,1,0.1))
legend(
  'bottomright',
  bty = 'n',
  col = colvector,
  pch = 15,
  row.names(pc_swiss_loadings)
)
par(opar)

# generate a biplot for each pair of important PCs (and show them on the same page)
#   note: the option choices is used to select the PCs - default is 1:2
opar <- par(no.readonly = TRUE)
par(mfrow = c(2,2))
biplot(
  pc_swiss,
  scale = 0,
  col = c('grey40','orange')
)
biplot(
  pc_swiss,
  choices = c(1,3),
  scale = 0,
  col = c('grey40','orange')
)
biplot(
  pc_swiss,
  choices = c(2,3),
  scale = 0,
  col = c('grey40','orange')
)
par(opar)
```


