---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706
### Lab 02 - Introduction to R for ML
#### Step by step tutorial
*Alessandro Pandini*  
Revision: 1.01

Contents adapted from:  
<https://github.com/dataspelunking/MLwR>

Reference:  
Chapter 6: Forecasting Numeric Data - Regression Methods  
Machine Learning with R - Second Edition  
Brett Lantz  
Packt Publishing 2015

***
#### Table of contents

R packages  
1. [Install and use R packages](#Install_and_use_R_packages)  
Example of ML data analysis  
2. [Data exploration (and preparation)](#Data_exploration)  
3. [Model training](#Model_training)  
4. [Model evaluation](#Model_evaluation)  
5. [Prediction](#Prediction)  
Writing R scripts  
6. [Saving plots](#Saving_plots)  
7. [R functions](#R_functions)  

***

#### 1. Install and use R packages{#Install_and_use_R_packages}  
```{r}
#   relevant websites:
#       https://cran.r-project.org
#       https://cran.r-project.org/web/packages
#       https://cran.r-project.org/web/views
#
#   relevant website for Machine Learning:
#       https://cran.r-project.org/web/views/MachineLearning.html

# list currently installed packages
#   note: the concept of libraries and packages in R
library()
# a more detailed list of currently installed packages
in_packages <- installed.packages()
head(in_packages)
colnames(in_packages)

# try to load a package (this may return an error!)
library(dplyr)

# install the missing package from CRAN
#   note: see also the RStudio menu Tools > Install Packages...
#         see also the RStudio Packages tab
#         packages can also be installed fomr .tgz files
install.packages("dplyr", repos = "http://cran.us.r-project.org")

# try to load the library now
library(dplyr)

# test it on the built-in iris dataset...
head(iris)

# example of select function from dplyr package
select(iris, Petal.Length)

# alternative way to access a package function (w/o loading the package)
dplyr::select(iris, Petal.Length)

# execute the examples from the help
?mean
example(mean)
?select
example(select)

# load the country demographics, economics and region datasets
ddemo <- read.csv("country_demo.csv", stringsAsFactors = F)
decon <- read.csv("country_econ.csv", stringsAsFactors = F)
dregion <- read.csv("country_region.csv", stringsAsFactors = F)

# merge datasets by country and year using inner_join
ij_data <- inner_join(ddemo, decon, by = c("country", "year"))

# merge datasets by country and year using left_join
lj_data <- left_join(ddemo, decon, by = c("country", "year"))

# compare the results of the two join commands
#   note: for a visual explanation of dplyr join functions see
#       https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf
ij_data
lj_data

# merge the dregion into the ij_data dataset
ijr_data <- inner_join(ij_data, dregion, by = "country")

# inspect the resulting dataset
#   note: region variable values have been repeated
ijr_data


# for more information on dplyr for table / data manipulation:
#       https://cran.rstudio.com/web/packages/dplyr/vignettes/dplyr.html

# install package psych for later use
if(require(psych) == FALSE){
    install.packages("psych", repos = "http://cran.us.r-project.org")
}
```

#### 2. Data exploration and preparation{#Data_exploration}  
```{r}
# read the data from a .csv file
insurance <- read.csv("insurance_data.csv", stringsAsFactors = TRUE)

# inspect the data
str(insurance)

# summarize the charges variable, i.e. the dependent variable
summary(insurance$expenses)

# plot a histogram of insurance charges
hist(insurance$expenses)

# get help on the hist command
?hist

# plot a customised histogram of insurance charges
hist(
    insurance$expenses,
    breaks = seq(0, 65000, 1000),
    xlab = 'Expenses in $1,000',
    ylab = 'Frequency',
    main = 'Insurance data: charges',
    col = rgb(0, 1, 0),
    border = 'white',
    xaxt = 'n'
)
axis(1, at = seq(0, 65000, 5000), seq(0, 65, 5))
abline(v = median(insurance$expenses), lty = 2)
legend('topright', 'median expenses', lty = 2, bty = 'n')

# plot expenses distribution by group of categorical variables - boxplot
boxplot(expenses ~ sex, data = insurance)
boxplot(expenses ~ smoker, data = insurance)
boxplot(expenses ~ region, data = insurance)

# generate frequency tables for each categorical variable
insurance_cat_table <- apply(insurance[, c("sex", "smoker", "region")], 2, table)

# visualize the tables
insurance_cat_table
insurance_cat_table$region

# calculate the percentage of non/smokers
insurance_smoker_perc <- insurance_cat_table$smoker / sum(insurance_cat_table$smoker) * 100
insurance_smoker_perc

# create label vectors for numerical and categorical variables
insurance_cat_names <- c("sex", "smoker", "region")
insurance_num_names <- c("age", "bmi", "children", "expenses")

# exploring relationships among features: correlation matrix
insurance_num_cor <- cor(insurance[insurance_num_names])

# visualize the correlation matrix
insurance_num_cor

# plot the relationships among features - scatterplot matrix
pairs(insurance[insurance_num_names])

# plot a more informative scatterplot matrix
psych::pairs.panels(insurance[insurance_num_names])
```

#### 3. Model training{#Model_training}  
```{r}
# create a linear model with lm from the stats package
ins_model <- lm(expenses ~ age + children + bmi + sex + smoker + region, data = insurance)

# equivalent lm call with short notation for the formula
ins_model <- lm(expenses ~ ., data = insurance)

# inspect the model
ins_model
# get attributes of the model
attributes(ins_model)
ins_model$coefficients
```

#### 4. Model evaluation{#Model_evaluation}  
```{r}
# get summary statistics on the model
ins_model_summary <- summary(ins_model)
ins_model_summary
# get attributes of the summary
attributes(ins_model_summary)
ins_model_summary$coefficients
ins_model_summary$adj.r.squared

# diagnostics plots for lm model
#   note: the generic call plot on an lm object is mapped to the plot.lm function
#         see help(plot.lm) for details
#   record current graph parameters
opar <- par(no.readonly = TRUE)
#   set plot for for panels
par(mfrow = c(2,2))
plot(ins_model)
#   reset graph parameters
par(opar)
```

#### 5. Prediction{#Prediction}  
```{r}
# read testset csv file
insurance_test <- read.csv("insurance_test_data.csv", stringsAsFactors = TRUE)

# filter for independent variables
insurance_indep_names <- c("age", "sex", "bmi", "children", "smoker", "region" )
insurance_test_indep <- insurance_test[,insurance_indep_names]
insurance_pred <- predict(ins_model, insurance_test_indep)

# check correlation between predicted and actual
cor(insurance_pred, insurance_test$expenses)

# plot predicted vs actual
#   record current graph parameters
opar <- par(no.readonly = TRUE)
par(pty = 's')
plot(
    insurance_pred ~ insurance_test$expenses,
    ylab = 'predicted',
    xlab = 'actual',
    main = 'expenses in $1000',
    xaxt = 'n',
    yaxt = 'n',
    xlim = c(0, 65000),
    ylim = c(0, 65000)
)
axis(1, at = seq(0, 65000, 5000), seq(0, 65, 5))
axis(2, at = seq(0, 65000, 5000), seq(0, 65, 5))
abline(a = 0, b = 1, col = 'red', lty = 2)
#   reset graph parameters
par(opar)
```

#### 6. Saving plots{#Saving_plots}  
```{r}
# save a plot to a png file
#   open the device png
png(file = "insurance_expenses_hist_simple.png")
#   plot the graph
hist(insurance$expenses)
#   close the device
dev.off()

# save a plot to a pdf file
#   note: EPS, BMP, JPEG and TIFF are also supported
pdf(file = "insurance_expenses_hist_simple.pdf")
#   plot the graph
hist(insurance$expenses)
#   close the device
dev.off()

# save a plot for a more complex example
png(file = "insurance_expenses_hist.png", width = 640, height = 480)
# plot the graph
hist(
    insurance$expenses,
    breaks = seq(0, 65000, 1000),
    xlab = 'Expenses in $1,000',
    ylab = 'Frequency',
    main = 'Insurance data: charges',
    col = rgb(0, 1, 0),
    border = 'white',
    xaxt = 'n'
)
axis(1, at = seq(0, 65000, 5000), seq(0, 65, 5))
abline(v = median(insurance$expenses), lty = 2)
legend('topright', 'median expenses', lty = 2, bty = 'n')
# close the device
dev.off()
```

#### 7. R functions{#R_functions}  
```{r}
# simple example to calculate the median of different features
colMedian <- function(dframe){
   medianVector <- apply(dframe, 2, median)
   return(medianVector)
}
#   inspect the function
colMedian
#   run the function on a dataframe
colMedian(insurance[insurance_num_names])

# simple example of a plotting function
colorHist <- function(dvector, dcolor){
    hist(
        dvector,
        col = dcolor,
        border = 'white'
    )
}
#   inspect the function
colorHist
#   run the function on a dataframe
colorHist(insurance$expenses, 'blue')
```
