---
title: "R Notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706
### Lab 06 - Prediction methods in R
#### Example of use of Self Organising Maps
*Alessandro Pandini*   
Revision: 1.01

***
#### Table of contents

1. [Install kohonen package](#Install_kohonen_package)
2. [Data preparation](#Data_preparation)
3. [SOM training](#SOM_training)
4. [SOM visualisation](#SOM_visualisation)

***

#### 1. Install kohonen package}{#Install_kohonen_package}
```{r}

# install and load the kohonen package from CRAN
if(require(kohonen) == FALSE){
  install.packages('kohonen')
}
```

#### 2. Data preparation{#Data_preparation}
```{r}

# read the data from the olive_oil.csv file
olive_oil <- read.csv("olive_oil.csv")

# read the geographical labels for the data
#   note: for this example the column region from this dataset will
#     be used as target variable for training
olive_oil_region <- read.csv("olive_oil_region.csv")

# inspect the data
str(olive_oil)

# the data frame needs to be 'cast' into a matrix for som training
#   note: R has an as.matrix function for this purpose
olive_oil_matrix <- as.matrix(olive_oil)

# the label vector needs to be 'cast' into a factor for som training
olive_oil_region_label <- as.factor(olive_oil_region$region)
```

#### 3. SOM training{#SOM_training}

```{r}

# set the random seed
set.seed(1999)

# train SOM model with 10x10 hexagonal grid
#   note: kohonen package provides different functions for training
#     som for unsupervised learning
#     xyf for supervised learning
olive_oil_SOM <- xyf(olive_oil_matrix, olive_oil_region_label,
                     grid = somgrid(10, 10, 'hexagonal'))
```

#### 4. SOM visualisation{#SOM_visualisation}
```{r}

# set color palettes for map plotting
#   note: the return value of colorRampPalette is a function that can
#     be used to generate a set of colours in the range of the palette
#       e.g. ncounts_palette(5) will return five colours
ncounts_palette <- colorRampPalette(c("white", "blue"), space = "rgb")
quality_palette <- colorRampPalette(c("white", "red"), space = "rgb")

# plot the number of objects mapped to each neuron
plot(
  olive_oil_SOM,
  shape = 'straight',
  type = 'counts',
  palette.name = ncounts_palette,
  main = "Number of objects per neuron"
)

# plot the mean distance of objects mapped to each neuron
plot(
  olive_oil_SOM,
  shape = 'straight',
  type = 'quality',
  palette.name = quality_palette,
  main = "Mean distance of objects in each neuron"
)

# plot the values of variables describing each neuron
#   for the input variables (from the olive_oil_matrix)
#   for the output variable (from olive_oil_region$region)
par(mfrow = c(1,2))
plot(
  olive_oil_SOM,
  shape = 'straight',
  type = 'codes',
  palette.name = rainbow,
  main = c("Fatty acid", "Region of origin")
)
par(mfrow = c(1,1))
```
