---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706
### Lab 06 - Prediction methods in R
#### Tutorial
*Alessandro Pandini*  
Revision: 1.01

***
#### Table of contents

1. [Install neural network package](#Install_neural_network_package)
2. [Data preparation](#Data_preparation)
3. [Neural network training](#Neural_network_training)
4. [Neural network prediction](#Neural_network_prediction)

***

#### 1. Install neural network package{#Install_neural_network_package}
```{r}
# install and load the neuralnet package from CRAN
if(require(neuralnet) == FALSE){
  install.packages('neuralnet')
}
```

#### 2. Data preparation{#Data_preparation}
```{r}

# read the data from the concrete.csv file
concrete <- read.csv("concrete.csv")

# inspect the data
str(concrete)

# transform the data using a min-max function
#   note: this will make the data more suitable for use with NN
#     as the attribute values will be on a narrow interval around zero
# first define a MinMax function
MinMax <- function(x){
  tx <- (x - min(x)) / (max(x) - min(x))
  return(tx)
}
# then apply the function to each column of the data set
#   note: the apply function returns a matrix
concrete_minmax <- apply(concrete, 2, MinMax)

# the matrix needs to be 'cast' into a data frame
#   note: R has an as.data.frame function for this purpose
concrete_minmax <- as.data.frame(concrete_minmax)

# create a 70/30 training/test set split
n_rows <- nrow(concrete_minmax)
# sample 70% (n_rows * 0.7) indices in the ranges 1:nrows
training_idx <- sample(n_rows, n_rows * 0.7)
# filter the data frame with the training indices (and the complement)
training_concrete_minmax <- concrete_minmax[training_idx,]
test_concrete_minmax <- concrete_minmax[-training_idx,]
```

#### 3. Neural network training{#Neural_network_training}
```{r}
# define a formula for predicting strength
concrete_formula = strength ~ cement + slag + ash + water + superplastic + coarseagg + fineagg + age

# train a neural network with 1 hidden node
concrete_nn_1 <- neuralnet(concrete_formula, data = training_concrete_minmax)

# train a neural network with 5 nodes on one hidden layer
#   note: the number of layers is set with the hidden option parameter
concrete_nn_5 <- neuralnet(concrete_formula, hidden = 5, data = training_concrete_minmax)

# train a neural network with 5 nodes on each of two hidden layers
concrete_nn_55 <- neuralnet(concrete_formula, hidden = c(5,5), data = training_concrete_minmax)

# plot the three neural networks and compare their structure
plot(concrete_nn_1)
plot(concrete_nn_5)
plot(concrete_nn_55)
```

#### 4. Neural network prediction{#Neural_network_prediction}
```{r}
# compute the prediction for each neural network
#   note: the strength attribute (column 9) is excluded from the test data set
pred_concrete_nn_1 <- compute(concrete_nn_1, test_concrete_minmax[,-9])
pred_concrete_nn_5 <- compute(concrete_nn_5, test_concrete_minmax[,-9])
pred_concrete_nn_55 <- compute(concrete_nn_55, test_concrete_minmax[,-9])

# create a table with actual values and the three predictions
#   note: predicted values are stored as net_result attribute of the prediction object
concrete_results <- data.frame(
  actual = test_concrete_minmax$strength,
  nn_1 = pred_concrete_nn_1$net.result,
  nn_5 = pred_concrete_nn_5$net.result,
  nn_55 = pred_concrete_nn_55$net.result
)

# calculate the correlation between actual and predicted values to identify the best predictor
cor(concrete_results[,'actual'], concrete_results[,c("nn_1","nn_5", "nn_55")])

# plot actual vs predicted values for the worst (blue) and best predictor (orange)
#   note: points is used to add points on a graph
plot(
  concrete_results$actual,
  concrete_results$nn_1,
  col = 'blue',
  xlab = 'actual strength',
  ylab = 'predicted strength',
  xlim = c(0,1),
  ylim = c(0,1)
)
points(
  concrete_results$actual,
  concrete_results$nn_55,
  col = 'orange'
)
abline(a = 0, b = 1, col = 'red', lty = 'dashed')
legend(
  'topleft',
  c('nn_1', 'nn_55'),
  pch = 1,
  col = c('blue', 'orange'),
  bty = 'n'
)
```

