---
title: "R Notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = '>')
```

### CS5706 
### Lab 01 - Introduction to R for ML  
#### Step by step tutorial  
*Alessandro Pandini*  
Revision: 1.01

Contents adapted from:  
<https://github.com/dataspelunking/MLwR>

Reference:  
Chapter 2: Managing and Understanding Data  
Machine Learning with R - Second Edition  
Brett Lantz  
Packt Publishing 2015  

***
#### Table of contents

R data structures  
1. [Vectors](#Vectors)  
2. [Factors](#Factors)  
3. [Matrices and Arrays](#Matrices_and_Arrays)  
4. [Lists](#Lists)  
5. [Data Frames](#Data_Frames)  

Managing data in R  
6. [Write and read data](#Write_and_read_data)  
7. [Basic data exploration](#Basic_data_exploration)  
8. [Basic data integration](#Basic_data_integration)  

***  

#### 1. Vectors {#Vectors}
```{r}
#   1 dimension
#   ordered elements of the same type:
#       integer     (numerical)
#       double      (numerical)
#       character
#       logical

# create vectors of data for three medical patients
#   note: the <- assignment operator in R
#         the c() combine operator in R
subject_name <- c("John Doe", "Jane Doe", "Steve Graves")
temperature <- c(98.1, 98.6, 101.4)
flu_status <- c(FALSE, FALSE, TRUE)

# inspect vector type
#   note: the typeof function in R
typeof(subject_name)
typeof(temperature)
typeof(flu_status)

# get help on a function in R
help(typeof)
?typeof

# access the second element in body temperature vector
temperature[2]

# access items in the range 2 to 3
temperature[2:3]

# access all elements except item 2 using the minus sign
temperature[-2]

# perform conditional selection using a logical vector
temperature[c(TRUE, TRUE, FALSE)]

# perform conditional selection using a logical vector
temperature[flu_status]

# perform conditional selection using a conditional statement
flu_status[temperature > 98.5]

# perform conditional selection using a condition statement
#   note: boolean operator
subject_name[temperature > 98.5 & flu_status == TRUE]

# include missing values (i.e. Not Available)
patient_name <- c("Jennifer Doe", "Bob Doe", "Steve Graves", "John Smith")
age <- c(56, 77, 101, NA)

# inspect vector
patient_name
age

# NULL object in R
vaccination <- NULL
vaccination
```

#### 2. Factors {#Factors}
```{r}
#   1 dimension
#   ordered elements of the same type for categorical / nominal values

# add gender factor
gender <- factor(c("MALE", "FEMALE", "MALE"))
gender

# add blood type factor
blood <- factor(c("O", "AB", "A"),
                levels = c("A", "B", "AB", "O"))
blood

# add ordered factor
symptoms <- factor(c("SEVERE", "MILD", "MODERATE"),
                   levels = c("MILD", "MODERATE", "SEVERE"),
                   ordered = TRUE)
symptoms

# check for symptoms greater than moderate
symptoms > "MODERATE"
```

#### 3. Matrices and Arrays {#Matrices_and_Arrays}  
```{r}
#   2 and n dimensions
#   ordered elements of the same type

# create a 2x2 matrix
m <- matrix(c(1, 2, 3, 4), nrow = 2)
m

# equivalent to the above
m <- matrix(c(1, 2, 3, 4), ncol = 2)
m

# create a 2x3 matrix
m <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
m

# create a 3x2 matrix
m <- matrix(c(1, 2, 3, 4, 5, 6), ncol = 2)
m

# extract values from matrixes
m[1, 1]
m[3, 2]

# extract rows
m[1, ]

# extract columns
m[, 1]

# create a 3D array
a <- array(c(1:24), dim = c(2,3,4))
a
```


#### 4. Lists {#Lists}  
```{r}
#   1 dimension
#   ordered elements of the different type (with named components)

# display information for a patient
subject_name[1]
temperature[1]
flu_status[1]
gender[1]
blood[1]
symptoms[1]

# create a list for the patient data
subject1 <- list(fullname = subject_name[1],
                 temperature = temperature[1],
                 flu_status = flu_status[1],
                 gender = gender[1],
                 blood = blood[1],
                 symptoms = symptoms[1])

# display the patient data
subject1

# get a single list value by position (returns a sub-list)
subject1[2]
typeof(subject1[2])
# checking the data type with the class() function
class(subject1[2])

# get a single list value by position (returns a numeric vector)
subject1[[2]]
typeof(subject1[[2]])

# get a single list value by name
subject1$temperature

# get several list items by specifying a vector of names
subject1[c("temperature", "flu_status")]

# access a list like a vector
subject1[2:3]

# create a list of lists
subject2 <- list(fullname = subject_name[2],
                 temperature = temperature[2],
                 flu_status = flu_status[2],
                 gender = gender[2],
                 blood = blood[2],
                 symptoms = symptoms[2])

subject3 <- list(fullname = subject_name[3],
                 temperature = temperature[3],
                 flu_status = flu_status[3],
                 gender = gender[3],
                 blood = blood[3],
                 symptoms = symptoms[3])

cohort <- list(subject1, subject2, subject3)

# inspect the list of lists
cohort

# access subject 2 data
cohort[[2]]
```

#### 5. Data Frames {#Data_Frames}  
```{r}
# create a data frame from medical patient data
pt_data <- data.frame(subject_name, temperature, flu_status, gender,
                      blood, symptoms, stringsAsFactors = FALSE)

# display the data frame
pt_data

# get a single column from a data frame
pt_data$subject_name

# get several columns by specifying a vector of names
pt_data[c("temperature", "flu_status")]

# this is the same as above, extracting temperature and flu_status
pt_data[2:3]

# accessing by row and column
pt_data[1, 2]

# accessing several rows and several columns using vectors
pt_data[c(1, 3), c(2, 4)]

# extract only column 1 for all rows
pt_data[, 1]
# extract only row 1 for all columns
pt_data[1, ]
# extract all rows and all columns
pt_data[ , ]

# other examples
pt_data[c(1, 3), c("temperature", "gender")]
pt_data[-2, c(-1, -3, -5, -6)]

# conditional selection
pt_data[pt_data$gender == 'MALE', ]
pt_data[pt_data$temperature < 100, ]
```

#### 6. Write and read data {#Write_and_read_data}  
```{r}
# write a data frame in a text file
write.table(pt_data, "pt_data.dat")
# write a data frame in a csv file
write.csv(pt_data, "pt_data.csv")

# save a data frame in an R binary file
save(pt_data, file = "pt_data.RData")

# show all data structures in memory
ls()

# remove the m and subject1 objects
rm(m, subject1)
ls()

# remove all objects
rm(list=ls())
ls()

# read the patient data from the text file
patient_data <- read.table("pt_data.dat", header = TRUE, stringsAsFactors = FALSE)
patient_data
ls()

# load the patient data from the R binary file
load("pt_data.RData")
ls()
pt_data

# read used cars data from a csv file
usedcars <- read.csv("usedcars.csv")
ls()
```

#### 7. Basic data exploration {#Basic_data_exploration}  
```{r}
# show the used cars data
usedcars
# get data frame head (first six lines) and tail (last six lines) parts
head(usedcars)
tail(usedcars)

# get data frame size
nrow(usedcars)
ncol(usedcars)
dim(usedcars)

# get column names
names(usedcars)

# display internal structure of data frame object
str(usedcars)

# display internal structure of data frame object
#   note: the effect of stringsAsFactors = FALSE in reading the data
str(patient_data)

# get summary of mileage feature
summary(usedcars$mileage)

# get statitics on mileage feature
mean(usedcars$mileage)      # mean
median(usedcars$mileage)    # median
range(usedcars$mileage)     # range
var(usedcars$mileage)       # variance
sd(usedcars$mileage)        # standard deviation
IQR(usedcars$mileage)       # Inter Quartile Range
quantile(usedcars$mileage)  # quantiles

# apply a function to more columns
apply(usedcars[,c("year","price","mileage")], 2, summary)
apply(usedcars[,c("year","price","mileage")], 2, median)

# get a frequency table for the year variable
table(usedcars$year)

# assign the returned value to a variable
median_price <- median(usedcars$price)
median_price

# visualize feature distribution - boxplot
boxplot(usedcars$mileage)

# visualize feature distribution - histogram
hist(usedcars$mileage)

# visualize feature distribution by group - boxplot
boxplot(price ~ year, data = usedcars)
```

#### 8. Basic data integration {#Basic_data_integration}  
```{r}
# load the country demographics, economics and region datasets
ddemo <- read.csv("country_demo.csv", stringsAsFactors = F)
decon <- read.csv("country_econ.csv", stringsAsFactors = F)
dregion <- read.csv("country_region.csv", stringsAsFactors = F)

# inspect the datasets
str(ddemo)
str(decon)
str(dregion)

# merge datasets by country and year
m_data <- merge(ddemo, decon, by = c("country", "year"))

# visualise and inspect the merged dataset
#   note: only common observations between the datasets are included
m_data
str(m_data)

# merge datasets by country and year including all instances from ddemo
#   note: argument all.x, all.y and all can be used to include all instances
#     see ?merge for more details
mx_data <- merge(ddemo, decon, by = c("country", "year"), all.x = TRUE)

# visualise and inspect the merged dataset
#   note: NAs have been added for missing observations
mx_data
str(mx_data)

# merge the dregion into the mx_data dataset
mxr_data <- merge(mx_data, dregion, by = "country")

# visualise and inspect the merged dataset
#   note: region variable values have been repeated
mxr_data
str(mxr_data)
```
